---
title: "Australian Bird Baths"
format: html
editor: visual
author: 
  - name: Vinny Chloros
    email: vinny.chloros@snhu.edu
    affiliations: 
      - name: Southern New Hampshire University
format: html
editor: visual
date: 2/4/2025
date-modified: today
date-format: long
code-fold: show
theme: cyborg
---

```{r required packages}
#| message: false

library(tidyverse)
library(tidymodels)
library(kableExtra)
library(patchwork)
```


## Abstract

This bird bath data from Cleary et al, 2016 is a collaborative effort between researchers and citizen scientists to track activity in bird baths throughout Australia. Because of the arid conditions in most of the region, bird baths might prove to be a vital resource for birds who rely on them as a water source. 

## Introduction

Bird baths are a common outdoor fixture used to provide a small pool of water for birds (and other backyard visitors) to use. It's not uncommon to see birds using a bird bath, but the scientific understanding for how, when, and why they use baths, at the time of this study, was not fully understood. To remedy this, a group of researchers used citizen scientists (i.e. members of the public collaborating with professionals) to collect data on birds they observed using bird baths throughout different areas of Australia. The study was conceptualized by researchers at Deakin University and Griffith University, and the data was collected over two four-week periods: one in 2014 in winter, and the other in 2015 in summer. In the data set, observations are classified by the year they took place, whether the location was urban or rural based on human population size, what region it took place in (delineated by a mixture of location, climate, and other environmental factors), bird type (species), and bird count. 

```{r initial data}
#| message: false
tuesdata <- tidytuesdayR::tt_load('2021-08-31')

bird_baths <- tuesdata$bird_baths
```

```{r data split}
#| message: false
birds_splits <- initial_split(bird_baths, prop = 0.5)

exploratory_data <- training(birds_splits)
test_data <- testing(birds_splits)

```

## Question(s)

1.  Does bird bath usage vary by season and/or region?

2.  Are more birds spotted in urban or rural locations?

3.  How does climate factor into bird bath usage?

## Exploratory Data Analysis (EDA)

### Understanding Data Format

As a first step, let's take a quick peek into the data to see what our values look like at a glance.

```{r start exploring}
exploratory_data$survey_year <- as.character(exploratory_data$survey_year)

exploratory_data |>
  head() |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

At this point I noticed that many of the inputs for `bird_count` were 0. This struck me as odd. I tried a few instances of `count()` functions to get a feel for how prevalent birds were within different years and environments.

```{r counts}
exploratory_data |>
  count(survey_year) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  count(urban_rural) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  count(bioregions) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

Around this point, I tried to isolate only the instances with at least one bird sighting, since I wasn't sure of the significance of the observations with 0 for `bird_count`.

```{r filtered count}
exploratory_data |>
  filter(bird_count > 0) |>
  count(bioregions) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

I realized that this table wasn't actually giving me the number of birds seen in these regions, but instead just the number of instances that birds were found. Within `bird_count`, the actual number of birds seen could vary from 0 to \>100.

Here, I figured out how the data was actually sorted within the data frame: there is an instance for each possible bird species within the overarching habitats and dates. To find what I'm actually interested in, I'll have to summarize the number of birds observed within each specific instance (bioregion + urban/rural + year).

```{r summary}
exploratory_data |>
  filter(bird_count > 0) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  filter(bird_count > 0) |>
  group_by(urban_rural) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  filter(bird_count > 0) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

Now, I finally have the initial stats that I'm looking for! You can see the total number of birds found between the years surveyed, between rural and urban areas, and different regions.

Now, I'll combine some of these variables so we can see how bird spottings differ when we summarize both year *and* region.

```{r year and region}
exploratory_data |>
  filter(bird_count > 0) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
I can get more particular and filter out the results with only one bird sighting.

```{r more than one bird}
exploratory_data |>
  filter(bird_count > 1) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
This code came up with a huge list of regions and years that come up with NA. To try to see more, I filtered out values that returned NA...

```{r more birds no NA}
exploratory_data |>
  filter(bird_count > 1) |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

And there were no results! It seems like most instances of more than one bird sighting in the data aren't associated with most of the identifying information.

If I sort by `bird_type`, maybe we can find some of those lost values.

```{r lost birds}
exploratory_data |>
  filter(bird_count > 0) |>
  select(bird_type, survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
Here they are! It seems like there are many instances of birds spotted, but the regions and survey year were unknown. In further plots, since I'm focused on examining how region variables like region, urban status, and season/year, I'll be excluding observations that lack information for all three of them.

### Starting Plots, Single Variable by Count

Now, back to the investigation, I'll try to see if I can summarize the total birds spotted by the region of they were spotted in.

```{r regions}
#| message: false

exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion",
       x = "Number of Birds",
       y = "Region") +
  scale_fill_discrete() +
  guides(fill = "none")

EDA_bioregion

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(bioregions)) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")

```

```{r regions total}
#| echo: false
#| message: false
#| output: false

exploratory_data |>
  filter(!is.na(bioregions)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```
</br>
Success! From this graph, we can see that the largest amount of birds were seen in the Sydney Basin, which makes up around 37% of all the observations. In comparison, South Eastern Queensland (SE Queensland) has the next largest amount of sightings with just over half the amount of birds (~22% of the data). Brigalow Belt South and Victorian Volcano Plain were almost tied for the least amount of birds, only differing by 2 sightings (both around 2%).

```{r year}
exploratory_data |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Bird Count by Year",
       x = "Number of Birds",
       y = "Survey Year") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(survey_year)) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

```{r}
#| output: false
#| echo: false

exploratory_data |>
  filter(!is.na(survey_year)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```


Between the two years, 2014 saw more birds than 2015 by a slim margin. 2014 sightings represent 54% of the sightings while 2015 represents the other 46%. Despite the different seasons, it seems like the total amount of birds did not vary by much.

```{r urbanness}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = urban_rural, fill = urban_rural)) +
  labs(title = "Bird Count in Urban and Rural Areas",
       x = "Number of Birds",
       y = "") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#A2CD5A", "#CDAA7D"))

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(urban_rural)) |>
  group_by(urban_rural) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

##exploratory_data |>
##  filter(!is.na(urban_rural)) |>
##  summarize(sum(bird_count, na.rm = TRUE))
```

Most of the bird bath sightings were in urban regions, making up 60% of the total sightings, while rural sightings made up the other 40%. Based on this, there might be some evidence that urban areas experienced more sightings, but further analysis would be needed for further support.

### Further Analysis, Multiple Variables Against Count

```{r region and year}
exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion and Year",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(. ~ survey_year) +
  scale_fill_discrete() +
  guides(fill = "none")
```

By comparing bioregions by year, there is a clear difference in distribution. In 2014, in winter, there were only a 5 bioregions that observations took place. Sydney Basin observations made up the majority of the 2014 bird count with over 700 birds tracked (around half of all 2014 observations). In contrast, 2015 featured observations of birds from every bioregion. The Sydney Basin is accompanied by SE Queensland and SE Highlands as the most common bioregions, with around 200 observations each.


```{r region and urbanity}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion and Urbanness",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(. ~ urban_rural) +
  scale_fill_discrete() +
  guides(fill = "none")
```

Comparing rural to urban observations across bioregions shows that all bioregions are accounted for in both groups. Urban areas feature Sydney Basin as the most common region by far (just under 800 observations) compared to SE Queensland, the next most common, with only around 300 observations. This might be caused by the city of Sydney, Australia's most populous city, lying within the basin. This might support the idea the more birdwatchers around leads to more observations.

```{r urbanity and year}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Bird Count by Survey Year and Urbanness",
       x = "Number of Birds",
       y = "Survey Year") +
  facet_grid(. ~ urban_rural) +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E")) +
  guides(fill = "none")
```

Here, I compared urbanness to the survey year of observations. Rural areas had less total sightings than urban areas between both years. In rural areas saw more bird observations in summer than winter, but urban areas saw more observations in winter than in summer.

```{r all variables}
exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by All Variables",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(urban_rural ~ survey_year) +
  scale_fill_discrete() +
  guides(fill = "none")
```

This plot encompasses all of the variables used in analysis so far. The observations between both years have a similar distribution as the previous graph comparing bioregion and year: 2014 had birds only observed within 5 regions while 2015 had observations in every region. This is true in both rural and urban areas. Between the two, it again appears that urban areas have a larger proportion of bird observations in general

## Hypotheses

1. Bird bath usage is limited to a few bioregions in winter and varies more widely in summer due to a greater need across regions for water as becomes more scarce.

2. Urban areas have more observations than rural areas because of the number of birdwatchers available.

3. Arid regions like the Sydney Basin see more bird bath usage because of their dry conditions.

## Methodology

Now, I can use these hypotheses to analyze the full data set.

### Poking Around, Single Variables

```{r full data}
bird_baths$survey_year <- as.character(bird_baths$survey_year)

bird_baths |>
  head() |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

```{r full regions}
EDA_bioregion <- exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "EDA Bird Count by Bioregion",
       x = "Number of Birds",
       y = "") +
  theme(axis.text.y = element_blank()) +
  scale_fill_discrete() +
  guides(fill = "none")

full_bioregion <- bird_baths |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Total Bird Count by Bioregion",
       x = "Number of Birds",
       y = "Region") +
  scale_fill_discrete() +
  guides(fill = "none")

(full_bioregion + EDA_bioregion)

bird_baths |>
  filter(bird_count > 0) |>
  filter(!is.na(bioregions)) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```

```{r total region count}
#| echo: false
#| message: false
#| output: false

bird_baths |>
  filter(!is.na(bioregions)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```

The total distribution of birds by region is extremely similar to that of the EDA data. Sydney Basin is again the region with the largest sum of bird observations, making up about 35% of the data (very similar to the ~37% of the EDA data). SE Queensland is the runner-up once again with around 21% of the data, one point off from the previous ~22%.

```{r}
full_year <- bird_baths |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Total Bird Count by Year",
       x = "Number of Birds",
       y = "Survey Year") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

EDA_year <- exploratory_data |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "EDA Bird Count by Year",
       x = "Number of Birds",
       y = "") +
  theme(axis.text.y = element_blank()) +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

(full_year + EDA_year)

bird_baths |>
  filter(bird_count > 0) |>
  filter(!is.na(survey_year)) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

```{r}
#| output: false
#| echo: false

bird_baths |>
  filter(!is.na(survey_year)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```

Again, the difference between the number of observations in each year reflects what was observed within the EDA. More birds were observed in the winter of 2014 (~52%) than the summer of 2015 (~48%). Compared to the 54/46 split in the EDA, they are 
