---
title: "Australian Bird Baths"
format: html
editor: visual
author: 
  - name: Vinny Chloros
    email: vinny.chloros@snhu.edu
    affiliations: 
      - name: Southern New Hampshire University
date: 2/4/2025
date-modified: today
date-format: long
code-fold: show
bibliography: birdbath_references_biblatex.bib
csl: apa.csl
toc: true
theme: cyborg
---

```{r required packages}
#| message: false

library(tidyverse)
library(tidymodels)
library(kableExtra)
library(patchwork)
```


## Abstract

This bird bath data from Cleary et al, 2016 is a collaborative effort between researchers and citizen scientists to track activity in bird baths throughout Australia. Because of the arid conditions in most of the region, bird baths might prove to be a vital resource for birds who rely on them as a water source. 

## Introduction

Bird baths are a common outdoor fixture used to provide a small pool of water for birds (and other backyard visitors) to use. It's not uncommon to see birds using a bird bath, but the scientific understanding for how, when, and why they use baths, at the time of this study, was not fully understood. To remedy this, a group of researchers used citizen scientists (i.e. members of the public collaborating with professionals) to collect data on birds they observed using bird baths throughout different areas of Australia. The study was conceptualized by researchers at Deakin University and Griffith University, and the data was collected over two four-week periods: one in 2014 in winter, and the other in 2015 in summer. In the data set, observations are classified by the year they took place, whether the location was urban or rural based on human population size, what region it took place in (delineated by a mixture of location, climate, and other environmental factors), bird type (species), and bird count. 

```{r initial data}
#| message: false
tuesdata <- tidytuesdayR::tt_load('2021-08-31')

bird_baths <- tuesdata$bird_baths
```

```{r data split}
#| message: false

set.seed(411254)
birds_splits <- initial_split(bird_baths, prop = 0.5)

exploratory_data <- training(birds_splits)
test_data <- testing(birds_splits)

```

## Question(s)

1.  Does bird bath usage vary by season and/or region?

2.  Are more birds spotted in urban or rural locations?

1.  How does climate factor into bird bath usage?

## Exploratory Data Analysis (EDA)

### Understanding Data Format

As a first step, let's take a quick peek into the data to see what our values look like at a glance.

```{r start exploring}
exploratory_data$survey_year <- as.character(exploratory_data$survey_year)

exploratory_data |>
  head() |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

At this point I noticed that many of the inputs for `bird_count` were 0. This struck me as odd. I tried a few instances of `count()` functions to get a feel for how prevalent birds were within different years and environments.

```{r counts}
exploratory_data |>
  count(survey_year) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  count(urban_rural) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  count(bioregions) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

Around this point, I tried to isolate only the instances with at least one bird sighting, since I wasn't sure of the significance of the observations with 0 for `bird_count`.

```{r filtered count}
exploratory_data |>
  filter(bird_count > 0) |>
  count(bioregions) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

I realized that this table wasn't actually giving me the number of birds seen in these regions, but instead just the number of instances that birds were found. Within `bird_count`, the actual number of birds seen could vary from 0 to \>100.

Here, I figured out how the data was actually sorted within the data frame: there is an instance for each possible bird species within the overarching habitats and dates. To find what I'm actually interested in, I'll have to summarize the number of birds observed within each specific instance (bioregion + urban/rural + year).

```{r summary}
exploratory_data |>
  filter(bird_count > 0) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  filter(bird_count > 0) |>
  group_by(urban_rural) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")

exploratory_data |>
  filter(bird_count > 0) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

Now, I finally have the initial stats that I'm looking for! You can see the total number of birds found between the years surveyed, between rural and urban areas, and different regions.

Now, I'll combine some of these variables so we can see how bird spottings differ when we summarize both year *and* region.

```{r year and region}
exploratory_data |>
  filter(bird_count > 0) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
I can get more particular and filter out the results with only one bird sighting.

```{r more than one bird}
exploratory_data |>
  filter(bird_count > 1) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
This code came up with a huge list of regions and years that come up with NA. To try to see more, I filtered out values that returned NA...

```{r more birds no NA}
exploratory_data |>
  filter(bird_count > 1) |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  select(survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

And there were no results! It seems like most instances of more than one bird sighting in the data aren't associated with most of the identifying information.

If I sort by `bird_type`, maybe we can find some of those lost values.

```{r lost birds}
exploratory_data |>
  filter(bird_count > 0) |>
  select(bird_type, survey_year, bioregions, bird_count) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")
```
</br>
Here they are! It seems like there are many instances of birds spotted, but the regions and survey year were unknown. In further plots, since I'm focused on examining how region variables like region, urban status, and season/year, I'll be excluding observations that lack information for all three of them.

### Starting Plots, Single Variable by Count

Now, back to the investigation, I'll try to see if I can summarize the total birds spotted by the region of they were spotted in.

```{r regions}
#| message: false

exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion",
       x = "Number of Birds",
       y = "Region") +
  scale_fill_discrete() +
  guides(fill = "none")

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(bioregions)) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped") |>
  scroll_box(height = "300px")

```

```{r regions total}
#| echo: false
#| message: false
#| output: false

exploratory_data |>
  filter(!is.na(bioregions)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```
</br>
Success! From this graph, we can see that the largest amount of birds were seen in the Sydney Basin, which makes up around 37% of all the observations. In comparison, South Eastern Queensland (SE Queensland) has the next largest amount of sightings with just over half the amount of birds (~22% of the data). Brigalow Belt South and Victorian Volcano Plain were almost tied for the least amount of birds, only differing by 2 sightings (both around 2%).

```{r year}
exploratory_data |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Bird Count by Year",
       x = "Number of Birds",
       y = "Survey Year") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(survey_year)) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

```{r}
#| output: false
#| echo: false

exploratory_data |>
  filter(!is.na(survey_year)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```


Between the two years, 2014 saw more birds than 2015 by a slim margin. 2014 sightings represent 54% of the sightings while 2015 represents the other 46%. Despite the different seasons, it seems like the total amount of birds did not vary by much.

```{r urbanness}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = urban_rural, fill = urban_rural)) +
  labs(title = "Bird Count in Urban and Rural Areas",
       x = "Number of Birds",
       y = "") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#A2CD5A", "#CDAA7D"))

exploratory_data |>
  filter(bird_count > 0) |>
  filter(!is.na(urban_rural)) |>
  group_by(urban_rural) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

Most of the bird bath sightings were in urban regions, making up 60% of the total sightings, while rural sightings made up the other 40%. Based on this, there might be some evidence that urban areas experienced more sightings, but further analysis would be needed for further support.

### Further Analysis, Multiple Variables Against Count

```{r region and year}
exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion and Year",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(. ~ survey_year) +
  scale_fill_discrete() +
  guides(fill = "none")
```

By comparing bioregions by year, there is a clear difference in distribution. In 2014, in winter, there were only a 5 bioregions that observations took place. Sydney Basin observations made up the majority of the 2014 bird count with over 700 birds tracked (around half of all 2014 observations). In contrast, 2015 featured observations of birds from every bioregion. The Sydney Basin is accompanied by SE Queensland and SE Highlands as the most common bioregions, with around 200 observations each.


```{r region and urbanity}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion and Urbanness",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(. ~ urban_rural) +
  scale_fill_discrete() +
  guides(fill = "none")
```

Comparing rural to urban observations across bioregions shows that all bioregions are accounted for in both groups. Urban areas feature Sydney Basin as the most common region by far (just under 800 observations) compared to SE Queensland, the next most common, with only around 300 observations. This might be caused by the city of Sydney, Australia's most populous city, lying within the basin. This might support the idea the more birdwatchers around leads to more observations.

```{r urbanity and year}
exploratory_data |>
  filter(!is.na(urban_rural)) |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Bird Count by Survey Year and Urbanness",
       x = "Number of Birds",
       y = "Survey Year") +
  facet_grid(. ~ urban_rural) +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E")) +
  guides(fill = "none")
```

Here, I compared urbanness to the survey year of observations. Rural areas had less total sightings than urban areas between both years. In rural areas saw more bird observations in summer than winter, but urban areas saw more observations in winter than in summer.

```{r all variables}
exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by All Variables",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(urban_rural ~ survey_year) +
  scale_fill_discrete() +
  guides(fill = "none")
```

This plot encompasses all of the variables used in analysis so far. The observations between both years have a similar distribution as the previous graph comparing bioregion and year: 2014 had birds only observed within 5 regions while 2015 had observations in every region. This is true in both rural and urban areas. Between the two, it again appears that urban areas have a larger proportion of bird observations in general

## Hypotheses

1. Arid regions like the Sydney Basin see more bird bath usage because of their dry conditions.

2. Urban areas have more observations than rural areas because of the number of birdwatchers available.

3. Bird bath usage is limited to a few bioregions in winter and varies more widely in summer due to the migration of birds and their water usage varying by season.

## Methodology

Now, after getting a sense of the data's distributions, I can go one by one and see how my hypotheses compare to the full data set instead of the subset we've been using. A few instances of the full data set are shown below.

```{r full data}
bird_baths$survey_year <- as.character(bird_baths$survey_year)

bird_baths |>
  head() |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

### Birds by Region

```{r full regions}
EDA_bioregion <- exploratory_data |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "EDA Bird Count by Bioregion",
       x = "Number of Birds",
       y = "") +
  theme(axis.text.y = element_blank()) +
  scale_fill_discrete() +
  guides(fill = "none")

full_bioregion <- bird_baths |>
  filter(!is.na(bioregions)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Total Bird Count by Bioregion",
       x = "Number of Birds",
       y = "Region") +
  scale_fill_discrete() +
  guides(fill = "none")

(full_bioregion + EDA_bioregion)

bird_baths |>
  filter(bird_count > 0) |>
  filter(!is.na(bioregions)) |>
  group_by(bioregions) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = c("striped", "condensed")) |>
  row_spec(0, bold = T, color = "black") |>
  scroll_box(height = "300px")
```

```{r total region count}
#| echo: false
#| message: false
#| output: false

bird_baths |>
  filter(!is.na(bioregions)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```
</br>
To analyze how regions influenced observations, I created new plots based on the ones I used in my EDA. I used the full data set for each and applied filters that excluded the observations that lacked a value for bioregion or had a bird count of 0.

The total distribution of birds by region is extremely similar to that of the EDA data. Sydney Basin is again the region with the largest sum of bird observations, making up about 35% of the data (very similar to the ~37% of the EDA data). SE Queensland is the runner-up once again with around 21% of the data, one point off from the previous ~22%. SE Highlands and NSW North Coast are the next runners up, both with around 11% of the total observations.

To add a quantitative comparison of the data, I conducted a quick chi-squared test to see if the variation between bioregions was statistically significant. 

```{r}
clean_birdbaths <- bird_baths |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0)

regions_chi <- table(clean_birdbaths$bioregions)

regions_chi_result <- chisq.test(regions_chi)

regions_chi_result

regions_chi_result$p.value
```
I isolated the bioregions variable from the rest of the data set and created a table for the test to use containing only the bioregions and their individual bird counts. The test compares the given values to an even distribution (the expected value). My test returned with a p-test value of 0 meaning that the variation of bird counts between bioregions is statistically significant.

### Urban vs Rural Observations

```{r}
bird_baths |>
  filter(!is.na(urban_rural)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = urban_rural, fill = urban_rural)) +
  labs(title = "Bird Count in Urban and Rural Areas",
       x = "Number of Birds",
       y = "") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#A2CD5A", "#CDAA7D"))

bird_baths |>
  filter(bird_count > 0) |>
  filter(!is.na(urban_rural)) |>
  group_by(urban_rural) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```
</br>
Like bioregions, the plot of urban and rural observations is almost identical to the one made in the EDA. Urban observations make up about 60% of the observations while rural observations are about 40%. This perfectly matches the proportions seen in the EDA. 

```{r}
#| echo: false
#| message: false
#| output: false

exploratory_data |>
  filter(!is.na(urban_rural)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```

I can do another statistical test to see if this difference in proportions is significant using the number of urban observances compared to the total number of observances.

```{r}
urban_results <- prop.test(x = 3344, n = 5545, p = 0.5, correct = FALSE)

urban_results

urban_results$p.value
```
This proportion test does confirm that the proportion of urban observances is statistically significant with a p-value of about 3.57e-53.

### Seasonal Variability

```{r}
full_year <- bird_baths |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "Total Bird Count by Year",
       x = "Number of Birds",
       y = "Survey Year") +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

EDA_year <- exploratory_data |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = survey_year, fill = survey_year)) +
  labs(title = "EDA Bird Count by Year",
       x = "Number of Birds",
       y = "") +
  theme(axis.text.y = element_blank()) +
  guides(fill = "none") +
  scale_fill_manual(values = c("#8EE5EE", "#EEAD0E"))

(full_year + EDA_year)

bird_baths |>
  filter(bird_count > 0) |>
  filter(!is.na(survey_year)) |>
  group_by(survey_year) |>
  summarize(
    sum_bird_count <- sum(bird_count, na.rm = TRUE)
  ) |>
  kbl() |>
  kable_styling(bootstrap_options = "striped")
```

```{r}
#| output: false
#| echo: false

bird_baths |>
  filter(!is.na(survey_year)) |>
  summarize(sum(bird_count, na.rm = TRUE))
```

Again, the difference between the number of observations in each year reflects what was observed within the EDA. More birds were observed in the winter of 2014 (~52%) than the summer of 2015 (~48%). Compared to the 54/46 split in the EDA, they are almost identical. 

I conducted another proportion test to see if the variation of bird counts was statistically significant.

```{r}
year_results <- prop.test(x = 2904, n = 5545, p = 0.5, correct = FALSE)

year_results
```
The results show a p-value of 0.0004, meaning that the difference in bird counts between the two groups is statistically significant.

To go a step further, I wanted to reassess the previous plot that compares bird counts based on both bioregion and survey year by using the full data set.

```{r}
bird_baths |>
  filter(!is.na(bioregions)) |>
  filter(!is.na(survey_year)) |>
  filter(bird_count > 0) |>
  ggplot() +
  geom_col(aes(x = bird_count, y = bioregions, fill = bioregions)) +
  labs(title = "Bird Count by Bioregion and Year",
       x = "Number of Birds",
       y = "Region") +
  facet_grid(. ~ survey_year) +
  scale_fill_discrete() +
  guides(fill = "none")
```

Like the previous plots, this one closely resembles the plot made in the EDA. In 2014, winter in Australia, birds using baths were observed in only 5 bioregions: the Sydney Basin, SE Queensland, SE Highlands, SE Coastal Plain, and NSW South Western Slopes. In comparison, the observations in Australian summer are distributed throughout all the bioregions included in the study, and the observations were spread more evenly throughout the regions. 

## Conclusions

### Arid Regions

Both the EDA and the actual data analysis seemed to agree that total bird count varies significantly between different regions. My original hypothesis for this was that the aridity of regions contributed to high bird counts because birds would have less natural water sources and instead rely on bird baths placed by humans. To have a point of reference, I looked into data from the Interim Biogeographic Regionalisation for Australia (IBRA) report from the Australian government which designates the different bioregions of the country.

According to IBRA7, the latest version, the Sydney Basin is described as having "desiccated plateaus" and "skeletal soils, sands, and podzolics" [-@RefWorks:2023ibra]. Desiccated means dried or having had water removed while "skeletal" when used in terms of soil or other ground material means that its largely made of rock fragments [@RefWorks:2019skeletal]. Based on these factors, we can assume that the Sydney Basin is unlikely to have many available water features. Bird baths thus create a reliable source of water for many birds living in the region. 

Other areas with high bird counts differ from the Sydney Basin. The SE Highlands and the NSW North Coast are described as humid by IBRA but also host different woodlands, rainforests, or wetlands [-@RefWorks:2023ibra]. The higher temperatures of these areas might make bird baths convenient water sources, but they might play a smaller role for them than they do for birds in more arid areas.

### Urban Birders

My second hypothesis was that urban regions in the original data set would make up a larger proportion of observations since more human observers would be present to witness birds using bird baths. The analysis supported this hypothesis and showed that the difference in proportion between urban and rural observations was significant.

In addition to that metric, I wanted to include a visualization of Australia's bioregions and compare them to the most populated cities.

[![IBRA7 Bioregions [-@RefWorks:2023ibra]](ibra-regions.jpg)](https://www.dcceew.gov.au/sites/default/files/env/pages/5b3d2d31-2355-4b60-820c-e370572b2520/files/ibra-regions.pdf)

Here is the national division of Australia's bioregions according to the IBRA7. Some of the most populated cities are labeled in the image, like Sydney in it's eponymous basin and Brisbane in SE Queensland. 

In addition, I referenced figures based on data from iNaturalist and eBird, other large-scale collaborative projects dependent on individual observations, to compare how common bird observations were across different areas of Australia. Both maps were created from data collected by citizen scientists in Australia of birds (not specifically bird bath usage) between 2014 and 2015.

::: {layout-ncol=2}

![eBird Map [-@eBird]](eBirdMap.png)

![iNaturalist Map [-@iNat]](iNaturalistMap.png)

:::

The larger and darker circles on each map represent higher concentrations of observations. The largest cities in particular are visible on the iNaturalist map based on these circles, with the three largest ones on the east coast being Brisbane, Sydney, and Melbourne while following the coast clockwise. These locations correlate to large bird counts in their respective bioregions: SE Queensland for Brisbane, the Sydney Basin for Sydney, and, to a lesser extent, the SE Coastal Plain for Melbourne. eBird has a similar distribution, with most observations concentrated on the east coast of the country, but there are a greater number of large circles throughout the region. This may represent some of the less popular but still urban or suburban regions of the area and may contribute to the bird counts of regions like the SE Highlands that neighbor both Melbourne and Sydney.

Overall, there does seem to be a bias in observations in highly populated areas. Instead of deducing that most birds happen to be in areas with large human populations, the fact that these observations are from citizen scientists could suggest that a good part of these observations are incidental i.e. made when participants came across birds in day-to-day activities rather than seeking them out in more obscure locations. Therefore, using this data to make any broad conclusions about birds in more remote areas, like the desert-like regions of western side of the country, would be inappropriate. Recognizing these factors when choosing data is an important key to starting a robust analysis.

### Influence of Climate between Season and Bioregion

Between the two periods of observation, there were similar numbers of birds in both periods with a 52/48 split between the total bird count in each category. Although 2014 saw more birds using baths in total by a slim, but statistically significant, margin, the bigger difference lies in how bird count varies by region when splitting the counts by when they were taken. Observations from 2014 (in winter) were only occurring in 5 bioregions while observations in 2015 (in summer) occurred in all of the included bioregions. My initial hypothesis was that birds might be migrating throughout within the country to a few regions in the winter and then spread more evenly throughout the country in summer. While writing, I considered the fact that it was possible observations might be low for some areas since climate varied between different bioregions, meaning birds might have less use for an artificial water source. However, in that case, there would likely be at least a few sightings as birds might incidentally use bird baths anyway. It seems more reasonable that a complete lack of any bird bath usage indicates an absence of birds that instead resident in those bioregions at a different time of year. 

### Final Thoughts

This investigation using others' data was eye-opening and engaging. Based on this analysis alone, I could imagine a number of follow-up questions and investigations that could be started based on mine, and its clear how having data like this open to public use by others is beneficial; other scientist may think of use cases that the original researchers never considered and find a new way to contribute to the field's body of knowledge. 

## References

::: {#refs}
:::