---
title: "A Guide to Phylogenetic Analysis in R"
author: 
  - name: Vinny Chloros
    email: vinny.chloros@snhu.edu
    affiliations: 
      - name: Southern New Hampshire University
date: 4/8/2025
date-modified: today
date-format: long
code-fold: show
toc: true
csl: apa.csl
bibliography: phylo_refs.bib
format: html
editor: visual
---

```{r}
#| message: false

library(BiocManager)
library(ape)
library(msa)
library(phangorn)
library(ggtree)

```

## Introduction



## Sequence Alignment in MSA

When aligning sequences, there are a few things you might need to prep outside of R before you're ready to import your data. You'll need a set of sequences that you plan to draw a tree from, and as you're starting out, and you'll want to ensure that all your sequences are using the same gene. 

```{r}
felidaeSequences <- readAAStringSet("UnalignedFelidae.fas")

felidaeSequences
```
Here, you can see a subset of samples from the Felidae sequences. The first few sets of nucleotides are found under "seq" for each species. 

## Tree Construction with Phangorn



## Tree Visualization with ggtree



## References

::: {#refs}
:::